---
title: <span style="color:#FFD100">**Exploring Life Tables**<span>
output: 
  rmdformats::readthedown:
    css: custom.css
bibliography: pop_references.bib
---

# Introduction
Using IMHE and Census Data we complied a life table together in R for Palau in 2015. This was a very labour intensive process. We took our $_na_x$ values from Tawian in 1970, which we found to have a similar life expectancy as Palau. We got this data from the Human Mortality Database (HMD). We also decomposed the gender difference in life expectancy, saw the influence of heart failure on life expectancy between genders, and used the Sullivan method to estimate the years of healthy life left, with vision loss being our impairment.

# Loading the Packages

```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggthemes)
library(gt)
```

```{r load_data, message=FALSE}
df_main = read_csv("data/life_tables.csv")
df_female = df_main %>% filter(sex == "female")
df_male = df_main %>% filter(sex == "male")
```

# Initial data
Set up start of life table with the x, the n, the $_nN_x$ and the $_nD_x$

```{r set_up_columns}
life_table_female = tibble(
  x = df_female$x,
  # calculate interval and set open ended as NA
  n = c(diff(df_male$x), NA),
  nNx = df_female$pop,
  nDx = df_female$deaths
)

life_table_male = tibble(
  x = df_male$x,
  # calculate interval and set open ended as NA
  n = c(diff(df_male$x), NA),
  nNx = df_male$pop,
  nDx = df_male$deaths
)

```

## Female Table 2015

```{r female_table_initial}
life_table_female %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:4))
```

## Male Table 2015

```{r male_table_initial}
life_table_male %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:4))
```

# Calculating $_nm_x$
Simple division to calculate $_nm_x$

```{r nmx}
# calculate mortality/nmx
life_table_female$nmx = life_table_female$nDx /
  life_table_female$nNx

life_table_male$nmx = life_table_male$nDx /
  life_table_male$nNx
```

## Female Table 2015

```{r female_table_nmx}
life_table_female %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:5))
```

## Male Table 2015

```{r male_table_nmx}
life_table_male %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:5))
```

# Calculating $_na_x$
Use Taiwan 1970 Reference for the middle categories, Preston et.al for the 80+ category, and Coale and Demeny for 0 to 4 categories

```{r nax}
# calculate nax for 0 to 1
nax_natal_male = life_table_male$nmx[1] * 2.684 + 0.045
nax_natal_female = life_table_female$nmx[1] * 2.684 + 0.045

# calculate nax for 0 to 4
nax_toddler_male = -life_table_male$nmx[2] * 2.816 + 1.651
nax_toddler_female = -life_table_female$nmx[2] * 2.816 + 1.651

# calculate nax for 80+
nax_oa_male = 80 + 1/life_table_male$nmx[18] - 80/1-exp(-80*life_table_male$nmx[18])

nax_oa_female = 80 + 1/life_table_female$nmx[18] - 80/1-exp(-80*life_table_female$nmx[18])

# load taiwan 1970 nax
taiwan = read_csv("data/taiwan_nax.csv")

nax_male = c(nax_natal_male,
             nax_toddler_male,
             taiwan$male,
             nax_oa_male)

nax_female = c(nax_natal_female,
             nax_toddler_female,
             taiwan$female,
             nax_oa_female)

# join onto dataframe
life_table_male$nax = nax_male
life_table_female$nax = nax_female
```

## Female Table 2015

```{r female_table_nax}
life_table_female %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:6))
```

## Male Table 2015

```{r male_table_nax}
life_table_male %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:6))
```

# Calculating $_nq_a$ and $_np_x$

Calculate the chance of dying between intervals and the reverse (elsewise known as the chance of living!)
```{r nqa_npx}
# do female nqx
life_table_female$nqx = (life_table_female$n * life_table_female$nmx) / 
  (1 + (life_table_female$n - life_table_female$nax) * life_table_female$nmx)

# for 80 plus cat
life_table_female$nqx[18] = 1

# do male nqx
life_table_male$nqx = (life_table_male$n * life_table_male$nmx) / 
  (1 + (life_table_male$n - life_table_male$nax) * life_table_male$nmx)

# for 80 plus cat
life_table_male$nqx[18] = 1

# make numeric
life_table_female$nqx = life_table_female$nqx %>% as.numeric()
life_table_male$nqx = life_table_male$nqx %>% as.numeric()

# npx
life_table_female$npx = 1 - life_table_female$nqx
life_table_male$npx = 1 - life_table_male$nqx
```

## Female Table 2015

```{r female_table_}
life_table_female %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:8))
```

## Male Table 2015

```{r male_table_}
life_table_male %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:8))
```

# Calculating $l_x$ and $_nd_x$
Extrapolate upto a population of 100000 using a fun for loop, calculate deaths in this hypothetical population as such (though without the fun loop sadly)

```{r lx_ndx_nLx}
# do female
# initiate empty list with only 100000
lx_female_list = c(100000)

# use a for loop to 
for (i in seq(life_table_female$npx)) {
  lx = lx_female_list[i] * life_table_female$npx[i]
  lx_female_list = append(lx_female_list, lx)
}

# knock off zero last value
lx_female_list = head(lx_female_list, -1)

# do male
# initiate empty list with only 100000
lx_male_list = c(100000)

# use a for loop to 
for (i in seq(life_table_male$npx)) {
  lx = lx_male_list[i] * life_table_male$npx[i]
  lx_male_list = append(lx_male_list, lx)
}

# knock off zero last value
lx_male_list = head(lx_male_list, -1)

# bind together
life_table_female$lx = lx_female_list %>% as.numeric()
life_table_male$lx = lx_male_list %>% as.numeric()

# ndx, using the lead() function
life_table_female$ndx = life_table_female$lx - lead(life_table_female$lx,
                                                    default = 0)
life_table_male$ndx = life_table_male$lx - lead(life_table_male$lx,
                                                    default = 0)
```

## Female Table 2015

```{r female_table_lx}
life_table_female %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:10))
```

## Male Table 2015

```{r male_table_lx}
life_table_male %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:10))
```

# Calculating $_nL_x$
The hardest bit, a very complex for loop to calculate the Person Years in between intervals. Took me so so long to figure out.

```{r nLx}
# nLx
# do female
# initiate blank column
life_table_female$nLx = NA

# make a for loop, ignore warning as we will mannually fill in end value
for(i in seq(life_table_female$nLx)){
  if(i == 18){
    life_table_female$nLx[i] = life_table_female$lx[i] / life_table_female$nmx[i]
  } else {
  life_table_female$nLx[i] = (life_table_female$n[i] * life_table_female$lx[i + 1]) +
  (life_table_female$nax[i] * life_table_female$ndx[i])
}}

# nLx
# do male
# initiate blank column
life_table_male$nLx = NA

# make a for loop, ignore warning as we will mannually fill in end value
for(i in seq(life_table_male$nLx)){
  if(i == 18){
    life_table_male$nLx[i] = life_table_male$lx[i] / life_table_male$nmx[i]
  } else {
  life_table_male$nLx[i] = life_table_male$n[i] * life_table_male$lx[i + 1] +
  life_table_male$nax[i] * life_table_male$ndx[i]
}}
```

## Female Table 2015

```{r female_table_nLx}
life_table_female %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:11))
```

## Male Table 2015

```{r male_table_nLx}
life_table_male %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:11))
```

# Calculating $T_x$
Use reverse cumulative function on the reverse of Person-Years to get that adding backwards from 80 effect

```{r Tx}
# use reverse of the cumsum() function to calculate Tx
# doing this on the reverse of nLx
life_table_female$Tx = rev(cumsum(rev(life_table_female$nLx)))
life_table_male$Tx = rev(cumsum(rev(life_table_male$nLx)))
```

## Female Table 2015

```{r female_table_Tx}
life_table_female %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:12))
```

## Male Table 2015

```{r male_table_Tx}
life_table_male %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:12))
```

# Reaching 65
Calculate the probability of a 30 year old reaching 65 through multipling the probabilities of survival between 30 and 65

```{r reaching_sixty_five}
# use prod() function to multiply together npx values
prob_sixty_five_female = prod(life_table_female$npx[8:14])
prob_sixty_five_male = prod(life_table_male$npx[8:14])

# output resukts
paste0("Probability of Female Aged 30 reaching 65: ", 
       round(prob_sixty_five_female * 100, 2),"%")
paste0("Probability of Male Aged 30 reaching 65: ", 
       round(prob_sixty_five_male * 100, 2),"%")
```

# E Dagger/Life Expectancy
Calculate E dagger... exciting!!!

```{r life_exp}
# calculate e
life_table_female$ex = life_table_female$Tx / life_table_female$lx
life_table_male$ex = life_table_male$Tx / life_table_male$lx
```

## Female Table 2015

```{r female_table_e}
life_table_female %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:13))
```

## Male Table 2015

```{r male_table_e}
life_table_male %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:13))
```

Determine the Differences between Women and Men

```{r differences}
life_exp_female = life_table_female$ex[1]
life_exp_male = life_table_male$ex[1]

paste0("The Female Life Expectancy was found to be: ",
       round(life_exp_female, 2))
paste0("The Male Life Expectancy was found to be: ",
       round(life_exp_male, 2))
paste0("The Difference in Life Expectancy was found to be: ",
       round(life_exp_female - life_exp_male, 2))
```

# Decomposing Gender Gap
Set up decomposition table with data from the life table
```{r decomposing_set_up}
# set up dataframe
decomp = tibble(
  x = life_table_female$x,
  lx_f = life_table_female$lx,
  nLx_f = life_table_female$nLx,
  Tx_f = life_table_female$Tx,
  lx_m = life_table_male$lx,
  nLx_m = life_table_male$nLx,
  Tx_m = life_table_male$Tx
)

# show table
decomp %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:7))
```
## Calculating differnce
Try to get the fancy up arrow symbol in the dataframe. Use a very very tricky for loop to decompose between genders. Check at end if it works!

```{r decomp_calculate}
# create blank column
decomp$`n∆x` = NA

# use female a for loop to to the calc
for(i in seq(decomp$lx_m)){
  # open ended age group
  if(i == 18){
    decomp$`n∆x`[i] = (decomp$lx_m[i] / 100000) *
      ((decomp$Tx_f[i] / decomp$lx_f[i]) - 
         (decomp$Tx_m[i] / decomp$lx_m[i]))
  # other age groups
  } else {
    decomp$`n∆x`[i] = ((decomp$lx_m[i] / 100000) *
      ((decomp$nLx_f[i] / decomp$lx_f[i]) - 
         (decomp$nLx_m[i] / decomp$lx_m[i]))) +
      ((decomp$Tx_f[i + 1] / 100000) * ((
        decomp$lx_m[i] / decomp$lx_f[i]) -
          (decomp$lx_m[i + 1] / decomp$lx_f[i + 1])))
  }
}

# test decomposition accuracy
life_exp_diff_decomp = sum(decomp$`n∆x`)
paste0("Life Expectancy difference from Decomposition: ",
       round(life_exp_diff_decomp, 2))
paste0("Life Expectancy difference from Life Table: ",
       round(life_exp_female - life_exp_male, 2))
```

## Percentage contributions
Calculate the percentage contribution for each age group in the difference between men and women
```{r percentage_contribution}
decomp$percent = round(decomp$`n∆x` / sum(decomp$`n∆x`) * 100, 2)

# output table
decomp %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:9))
```

# Decomposing Cause of Death
More decomposition!!?? It's ok this is actually a bit easier I think

## Adding in Heart Failure Data and $_nm^i_x$

Where $_nm^i_x = \frac{_nd^i_x}{_nN_x}$

```{r  cause_of_death}
# add heart failure numbers and calculate nmix
decomp$nmix_f = df_female$heart_failure / df_female$pop
decomp$nmix_m = df_male$heart_failure / df_male$pop

# add base nmx
decomp$nmx_f = life_table_female$nmx
decomp$nmx_m = life_table_male$nmx

```

## Output Table

```{r output_decomp_md}
decomp %>%
  subset(select = c("x", "n∆x",	"nmix_f", "nmix_m",
                    "nmx_f",	"nmx_m")) %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:6))
```

# Calculating $_n∆^i_x$
Find the contribution in difference in age categories between men and women by Heart Failure as the cause of death. Does it help or hinder men's live expectancy when compared to women

```{r decomp_heart_failure}
# using formula from slides
decomp$`n∆ix` = (decomp$nmix_f - decomp$nmix_m) / 
  (decomp$nmx_f - decomp$nmx_m) *
  decomp$`n∆x`
```

## Output Table

```{r output_new_decomp}
decomp %>%
  subset(select = c("x","n∆x",	"nmix_f", "nmix_m",
                    "nmx_f",	"nmx_m",	"n∆ix")) %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:7))

```

```{r plot_decomp, warning = FALSE}
# create long data to plot
decomp$heart_failure = decomp$`n∆ix`
decomp$remaining = decomp$`n∆x` - decomp$`n∆ix`
decomp_toplot = decomp %>%
  select(x, remaining, heart_failure) %>%
  pivot_longer(cols = c(remaining, heart_failure),
               names_to = "type",
               values_to = "value") %>%
  mutate(type = ifelse(type == "heart_failure",
                       "Cardiovascular Disease",
                       "Remaining"))

# plot
p = ggplot(decomp_toplot, aes(x = x, y = value, fill = type)) +
  geom_col(stat = "identity", position = "stack", width = 5) +
  scale_x_continuous(breaks = seq(0, 85, by = 5),
                     labels = c("0",
                                      "1-4",
                                      "5-9",
                                      "10-14",
                                      "15-19",
                                      "20-24",
                                      "25-29",
                                      "30-34",
                                      "35-39",
                                      "40-44",
                                      "45-49",
                                      "50-54",
                                      "55-59",
                                      "60-64",
                                      "65-69",
                                      "70-74",
                                      "75-79",
                                      "80+")) +
  scale_y_continuous("Years of Female Life Surplus") +
  scale_fill_manual(values = c("Remaining" = "#AEDFF7",
                               "Cardiovascular Disease" = "#D17A7A")) +
  labs(x = "Age",
       y = "Years",
       title = "Life Expectancy Differences between Sex by Age Group and Cause of Death",
       subtitle = "Represents the cause-specific impact of Cardiovascular Disease",
       fill = "Contribution") +
  theme_solarized_2() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p

ggsave("stacked_bar_cause_specific.png", p)
```


# Sulivan Method for Impairment
Use the Sulivan Method to calculate life left without getting some form of vision loss

## Calculate $H_x$

```{r sulivan_one}
life_table_female$Hx = 1 - df_female$blindess / df_female$pop
life_table_male$Hx = 1 - df_male$blindess / df_male$pop
```

## Female Table 2015 (Partial or Total Loss of Vision)

```{r female_table_sul}
life_table_female %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:14))
```

## Male Table 2015 (Partial or Total Loss of Vision)

```{r male_table_sul}
life_table_male %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:14))
```

# Calculating $_nL^H_x$ and $T^H_x$

```{r nLHx_and_THx}
# nLHx
life_table_female$nLHx = life_table_female$nLx * life_table_female$Hx
life_table_male$nLHx = life_table_male$nLx * life_table_male$Hx

# THx
life_table_female$THx = rev(cumsum(rev(life_table_female$nLHx)))
life_table_male$THx = rev(cumsum(rev(life_table_male$nLHx)))
```

## Female Table 2015 (Partial or Total Loss of Vision)

```{r female_table_sula}
life_table_female %>%
  select(x, nNx, nDx, nLx, Tx, ex, Hx, nLHx, THx) %>% # cut so it fits
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:9))
```

## Male Table 2015 (Partial or Total Loss of Vision)

```{r male_table_sula}
life_table_male %>%
  select(x, nNx, nDx, nLx, Tx, ex, Hx, nLHx, THx) %>% # cut so it fits
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:9))
```


# Healthy vs Unhealthy Life Expectancy

Calculate E Dagger for Healthy and Unhealthy years left

```{r life_exp_health}
# eHx
life_table_female$eHx = life_table_female$THx /
  life_table_female$lx
life_table_male$eHx = life_table_male$THx / 
  life_table_male$lx

# eUNx
life_table_female$eUNx = life_table_female$ex -
  life_table_female$eHx
life_table_male$eUNx = life_table_male$ex -
  life_table_male$eHx
```

## Female Table 2015 (Partial or Total Loss of Vision)

```{r female_table_suli}
life_table_female %>%
  select(x, nLx, Tx, ex, Hx,
         nLHx, THx, eHx, eUNx) %>% # cut so it fits
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:9))
```

## Male Table 2015 (Partial or Total Loss of Vision)

```{r male_table_suli}
life_table_male %>%
  select(x, nLx, Tx, ex, Hx,
         nLHx, THx, eHx, eUNx) %>% # cut so it fits
  gt() %>%
  tab_style(style = cell_fill(color = "#1F78B4"),
            locations = cells_column_labels(columns = 1:9))
```

# [Return to Homepage](index.html)
